Experimental and numerical studies of the effects of resonant and random excitations of the proton beam in the LHC, with applications to the design of a pulsed hollow electron lens for active halo control

### 2018-02-20 resex.tex
Revised Section II on HEL

\subsection{General Overview}
\label{sec:hel:intro}

@@

The circulating beam (protons in the LHC case) is affected by the
electromagnetic field of the electron beam. For the application of
active halo control, the electron beam needs to generate an
electromagnetic field only at the location of the halo particles. This
field distribution can be achieved, for instance, by using a hollow
charge distribution in radius $r=\sqrt{x^2+y^2}$, uniformly
distributed between inner radius $R_1$ and outer radius $R_2$
(Fig.~\ref{hel_field}). In this case, the circulating proton beam
experiences the following radial kick $\theta(r)$:
%
\begin{equation}
  \label{eq:field_1}
  \theta(r)=\frac{f(r)}{(r/R_2)}\cdot \theta_{\rm max},
\end{equation}
%
where $f(r)$ is a shape function with
%
\begin{equation}
  \label{eq:field_2}
  f(r) =
  \begin{cases}
    0 &,\quad r< R_1,\\
    \frac{r^2-R_1^2}{R_2^2-R_1^2} &,\quad R_1 \leq r < R_2,\\
    1 &,\quad R_2 \leq r
  \end{cases}
\end{equation}
%
and $\theta_{\rm max}=\theta(R_2)$ is the maximum kick angle given by
%
\begin{equation}
  \label{hel_kick_max}
  \theta_{\rm max} = \theta(R_2) =
  \frac{2LI_T(1\pm\beta_e\beta_p)}{4\pi\epsilon_0  \cdot
    \left(p_0/q\right)_p \cdot \beta_e\beta_p
    c^2}\cdot\frac{1}{R_2},
\end{equation}
%
with $L$ the length of the HEL, $I_T$ the total electron beam current,
$\beta_{e}$ and $\beta_{p}$ the relativistic velocity parameters of
electrons and protons, $\left(p_0/q\right)_p = \left(B\rho\right)_p$
the magnetic rigidity for the proton beam reference particle, $c$ the
speed of light and $\epsilon_0$ the vacuum permittivity. The
$\pm$-sign in Eq.~\ref{hel_kick_max} represents the two cases of the
electron beam traveling in the direction of the proton beam
($v_e v_p>0$) leading to ``$-$" or in the opposite direction
($v_e v_p<0$) leading to ``$+$". For hollow electron beam collimation,
electrons and protons are chosen to counterrotate, so that the
magnetic and electric kicks add up. (For simplicity, the dependence of
the electron axial velocity on radius is neglected in
Eq.~\ref{hel_kick_max}.)

@@

For the HEL two modes of operation are currently under consideration,
the continuous (also referred to as `DC', or direct current, in this
paper) mode as standard operation mode and the pulsed mode. The main
benefit of pulsed HEL operation is the increase in halo removal rates
which might become essential as in ideal operation conditions
featuring low non-linearities, in particular low chromaticity and
octupole current, the DC mode is not
effective~\cite{hel_halo_hllhc_fitterer,hl_halo_ipac2017}.

For this purpose, two different pulsing patterns are thus considered
for the HL-LHC:

\begin{itemize}
	\item \textbf{random excitation:} The electron gun voltage is modulated between:
	\begin{eqnarray}
	U_{\mathrm{e-gun}}&=&a\cdot U_{\mathrm{max}}\\
	& &+(1-a)\cdot \mathrm{ran}(0,1)\cdot U_{\mathrm{max}}
	\end{eqnarray}
	with $U_{\mathrm{max}}$ the maximum voltage in DC operation, $a$ the modulation strength with $a\in[0,1]$, and $\mathrm{ran}(0,1)$ a uniformly distributed random number between~[0,1].
	\item \textbf{resonant excitation:} The HEL is switched on only every $k^{\mathrm{th}}$ turn. The excitation can then be represented by:
	\begin{eqnarray}\label{intro:eqn:1}
	f(t)&=&\sum_{n=-\infty}^{+\infty}\delta(t-n\cdot(kT)),
	\end{eqnarray}
	where $n$ is the turn number and $T$ the revolution time, and its Fourier series by:
	\begin{eqnarray}\label{intro:eqn:2}
	f(t)=\Sh_{kT}(t)
	&=&\frac{1}{kT}\sum_{n=-\infty}^{+\infty}e^{2\pi i f_nt} \\
	& & \quad \text{with} \ f_n=\frac{n}{k}f_{\rm rev},
	\end{eqnarray}
	where $\Sh_{kT}$ is the Dirac comb. As can be seen from the Fourier series, $k^{\mathrm{th}}$~turn pulsing in general drives $k^{\mathrm{th}}$ order resonances \cite{md_sim_hel_res_ex_fitterer}. Historically, the $k^{\mathrm{th}}$~turn pulsing was used in regular operation in the Tevatron for abort cleaning \cite{hel_tevatron_abortgap_zhang}.
\end{itemize}

As depicted in Fig.~\ref{fig:hel_field} and in evidence from Eqs.~\ref{eq:field_1}--\ref{eq:field_2}, the field at the beam core vanishes in case of an ideal HEL. Effects on the beam core thus only arise from imperfections, where possible sources are the bends of the HEL (see Section~\ref{core:sec:1}), as here the electron beam crosses directly the proton beam, and distortions in the electron beam profile (see Section~\ref{core:sec:2}). Both sources result in non-linear kicks~\cite{hel_bends_stancari,hel_model_polynomial_morozov}. In DC operation, these non-linear kicks stay in the shadow of the non-linearities otherwise present in the machine if the electron beam is kept at a decent quality. Tolerances on imperfections are therefore not particularly stringent and of no concern. The picture changes significantly in case of pulsed operation, in which case the electron gun voltage is modulated using a white or colored noise spectrum. If the electromagnetic field of the HEL does not vanish at the proton beam core, noise is transferred not only to the halo particles, as intended, but also to the beam core. Tolerances on the residual field in this case rapidly become much more stringent than in the case of DC operation. Studies of the effect of the HEL on the beam core therefore focus on this mode of operation which will also be the main subject of this paper.

%In comparison, the random excitation mode is the most effective and operationally easiest pulsing mode, but also the most stringent in terms of tolerances on the residual field at the proton beam center as it induces white noise exciting all beam frequencies. As the resonant excitation acts on specific resonances, the pulsing pattern has to be chosen according to the machine and beam parameters. On the one hand, this makes the operation more difficult as one has to find the most efficient $k^{\mathrm{th}}$~turn pulsing for each operation mode, on the other hand, it also relaxes the tolerances on the residual field at the beam core as, depending on the machine and beam configuration, it might be possible to find a resonant excitation pattern, which mainly affects the halo and not the beam core.

>>>>

\subsection{General overview}
\label{sec:hel:intro}

@@

The circulating beam (protons in the LHC case) is affected by the
electromagnetic field of the electron beam. For the application of
active halo control, the electron beam needs to generate an
electromagnetic field only at the location of the halo particles. This
field distribution can be achieved, for instance, by using a hollow
charge distribution in radius $r=\sqrt{x^2+y^2}$, uniformly
distributed between inner radius $R_1$ and outer radius $R_2$
(Fig.~\ref{fig:hel_field}). In this case, the circulating proton beam
experiences the following radial kick $\theta(r)$:
%
\begin{equation}
  \label{eq:field_1}
  \theta(r)=\frac{f(r)}{(r/R_2)}\cdot \theta_{\rm max},
\end{equation}
%
where $f(r)$ is a shape function with
%
\begin{equation}
  \label{eq:field_2}
  f(r) =
  \begin{cases}
    0 &,\quad r< R_1,\\
    \frac{r^2-R_1^2}{R_2^2-R_1^2} &,\quad R_1 \leq r < R_2,\\
    1 &,\quad R_2 \leq r
  \end{cases}
\end{equation}
%
and $\theta_{\rm max}=\theta(R_2)$ is the maximum kick angle given by
%
\begin{equation}
  \label{hel_kick_max}
  \theta_{\rm max} = \theta(R_2) =
  \frac{2LI_T(1\pm\beta_e\beta_p)}{4\pi\epsilon_0  \cdot
    \left(p_0/q\right)_p \cdot \beta_e\beta_p
    c^2}\cdot\frac{1}{R_2},
\end{equation}
%
with $L$ the length of the HEL, $I_T$ the total electron beam current,
$\beta_{e}$ and $\beta_{p}$ the relativistic velocity parameters of
electrons and protons, $\left(p_0/q\right)_p = \left(B\rho\right)_p$
the magnetic rigidity for the proton beam reference particle, $c$ the
speed of light and $\epsilon_0$ the vacuum permittivity. The
$\pm$-sign in Eq.~\ref{hel_kick_max} represents the two cases of the
electron beam traveling in the direction of the proton beam
($v_e v_p>0$) leading to ``$-$" or in the opposite direction
($v_e v_p<0$) leading to ``$+$". For hollow electron beam collimation,
electrons and protons are chosen to counterrotate, so that the
magnetic and electric kicks add up. (For simplicity, the dependence of
the electron axial velocity on radius is neglected in
Eq.~\ref{hel_kick_max}.)

@@

For the HEL, two modes of operation are currently under consideration:
the continuous mode (also referred to as `DC', or direct current, in
this paper) as standard operation mode, described above; and the
pulsed mode. The main benefit of pulsed HEL operation is the increase
in halo removal rates. A wider range of removal rates may become
important under operating conditions with small nonlinearities, in
particular low chromaticity and octupole current, when the DC mode may
be too slow~\cite{hel_halo_hllhc_fitterer, hl_halo_ipac2017}.

Two different pulsing patterns are considered for the HL-LHC. In both
cases, at each passage, a given bunch sees a different electron-lens
current and, therefore, experiences a different transverse kick. The
patterns are defined as follows:

\item \textbf{random excitation:} The extraction voltage in the
  electron gun is modulated according to the
  following expression:
%
  \begin{equation}
    U_{\mathrm{e-gun}} = (1-a) \cdot U_{\mathrm{max}} +
    a \cdot \eta \cdot U_{\mathrm{max}},
  \end{equation}
%
  where $U_{\mathrm{max}}$ is the maximum voltage, $a$ is the
  modulation strength, with $a \in [0,1]$, and $\eta$ is a uniformly
  distributed random number in the interval~$[0,1]$. Simulations and
  experiments, discussed below, were usually conducted with $a = 1$.

\item \textbf{resonant excitation:} The electron beam is switched on
  only every $k^{\mathrm{th}}$ turn. The excitation can be represented
  by the following expression:
  \begin{eqnarray}
    \label{intro:eqn:1}
    f(t) & = & \sum_{n=-\infty}^{+\infty}\delta\left[t-n\cdot(kT)\right],
  \end{eqnarray}
  where $n$ is the turn number and $T$ is the revolution period. Its
  Fourier representation is
  \begin{equation}
    \label{intro:eqn:2}
    f(t) = \Sh_{kT}(t) = \frac{1}{kT}\sum_{n=-\infty}^{+\infty}e^{2\pi i f_nt},
  \end{equation}
  with $f_n = n \cdot f_{\rm rev} / k$ and
  where $\Sh_{kT}$ is the Dirac comb. In general,
  $k^{\mathrm{th}}$~turn pulsing drives $k^{\mathrm{th}}$-order
  resonances~\cite{md_sim_hel_res_ex_fitterer}. This type of pulsing
  pattern was used in the Tevatron during regular collider operations
  for abort-gap cleaning~\cite{hel_tevatron_abortgap_zhang}.
\end{itemize}

For an axially symmetric electron lens, the field at the beam core
vanishes. Effects on the proton core arise from imperfections, with
two main sources: the injection and extraction bends of the HEL
(discussed below in Section~\ref{core:sec:1}), where the electron beam
crosses the proton beam; and distortions of the electron beam profile
during its propagation under magnetic focusing and space charge
(Section~\ref{core:sec:2}). Both sources result in nonlinear
kicks~\cite{hel_bends_stancari, hel_model_polynomial_morozov}. In
continuous operation, these nonlinear kicks are usually much smaller
than the machine nonlinearities. Tolerances on imperfections are
therefore not particularly stringent. The picture changes
significantly in case of pulsed operation. If the electromagnetic
field does not vanish at the proton beam core, noise or resonant kicks
are transferred not only to the halo particles, as intended, but also
to the beam core. Tolerances on the residual fields in this case
become much more stringent. Studies of the effects of the HEL on the
beam core therefore focus on this mode of operation, which is also the
main subject of this paper.
